Open archaeology, open source?

### 2023-06-21 analysis/_01-introduction.qmd
Some more work on the intro and background

Open archaeology constitutes a series of practices and principles in support greater transparency and access to research processes and outputs. The open science movement operates in various domains of research operations, such as publishing, data-sharing, and education, and continues to expand its reach into the arenas of bibliographic metadata, review processes, and grantsmanship. While these venues for transformation represent relatively novel ways of doing research, academic ventures into open source is more complicated. The open source movement has its own distinct history and developed out of pressures and circumstances beyond those in which the open science movement operates. However, the open source movement directly inspires many aspects of open science, and, crucially, informs how the open science movement has operationalized itself. For instance, the open science movement relies on infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plaintext, diffs, emphasis on formal licencing, the whole hacker aesthetic). But academic open source is distinct from open source that underlies massive projects that literally run the internet (and by extension the global economy), such as the Linux kernel, the Firefox web browser, and openSSL. While they may use similar tools and technical protocols, they are governed by different social and professional pressures. In other words, making your code open source has different meanings, impacts and implications for archaeologists and professional software developers.

This is also complicated by the kind of rhetoric about academic open source that, based on our personal experiences, is discordant with how academics actually _do_ open source. Is open source actually making research processes more transparent and improve research outcomes? Is it actually boosting efficiency by establishing a common store of knowledge and productive code? Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass? These outcomes are not a given. They only actually arise in contexts where participants adhere to and are motivated by warrants, professional norms, and governance strategies that encourage these results. Beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

@@

Our investigation considers how archaeological software development may be benefiting from or missing out on the affordances that open source development models-- specifically, the value added through working as part of a broader community of invested stakeholders, processes of iterative improvement, and increased code transparency.

>>>>

# Introduction

Open archaeology constitutes a series of practices and principles that support greater transparency to research processes and access to data that underlie synthetic outcomes (Beck and Neylon 2012: 479-480). As part of the broader open science movement, open archaeology promotes more thoughtful scholarly communication practices, most notably in operations relating to publishing, data-sharing, and education, but it continues to expand its reach into the arenas of bibliographic metadata, review processes, and grantsmanship. Archaeologists are also actively venturing into open source software development as means of sharing their research processes and creating tools and resources for general use.

However, academic implementation of open source software development models is complicated by the open source movement's distinct history and its nuanced relationship with the open science movement. The open source movement developed out of pressures and circumstances that differ from those in which the open science movement operates, and yet the former serves as direct inspiration for the latter. For instance, the Open Knowledge Foundation, which publishes a widely accepted defitnion of "open" in the context of scholarly communication, draws directly from the definition of "open source" published by the Open Source Initiative (@openknowledgefoundation2023; @opensourceinitiative2007). The open science movement further mimics open source by operationalizing scholarly communication through technical infrastructures and protocols that closely resemble systems and processes designed to develop open source software (e.g., the use of plaintext, diffs, emphasis on formal licencing, the whole hacker aesthetic). However, academic work, including the development of academic software, differs signififcantly from the work involved in massive open source projects that literally run the internet, such as the Linux kernel, the Firefox web browser, and openSSL. While they may use similar tools and technical protocols to manage coding operations, the open science and open source movements are governed by different social and professional pressures. In other words, publishing code openly on the web has different meanings, impacts and implications for archaeologists and professional software developers (Kelty 2008:xx).

The relationship between open science and open source is also complicated by rhetorical claims that, based on our personal experiences, are discordant with how academics actually _do_ open source. Does academic open source actually make research processes more transparent and improve research outcomes? Is it actually boosting efficiency by establishing a common store of knowledge and productive code? Is it actually helping to foster new globe-spanning connections and lead to novel research trajectories that would not otherwise come to pass? Despite hopeful aspirations espoused by open archaeology advocates (Kansa  et al. 2014; Kintigh et al. 2015; others), these outcomes are not a given. Based on casual observations from our own experiences, we find that these outcomes only actually arise in contexts where participants adhere to and are motivated by warrants, professional norms, and governance strategies that encourage these results. Unfortunately, these values are not widespread among archaeologists, even those who are staunch supporters of open science, due to systemic challenges and tensions.

That being said, archaeologists are prolific software developers, as is made evident by open-archaeo, a comprehensive list of open source archaeological software and resources, which records XXX items as of the time of publication. But beyond simply making their code available on the web, do archaeologists also implement social strategies to advance open source ideals? Does archaeological open source actually help achieve greater transparency, sustainability, and community participation? And if not, what does it actually achieve?

@@

Our statistical and network analyses consider how archaeological software development may be benefiting from or missing out on the affordances that open source development models provide, specifically the value added through working as part of a broader community of invested stakeholders, processes of iterative improvement, and increased code transparency. We found that the vast majority of projects lack any collective effort and die shortly after they are initiated. Projects that are maintained by multiple contributors tend to be backed by funded initiatives who employ locally and socially engaged individuals, or are warranted by participation in a scholarly community exhibiting genuine need for a particular kind of resource. Moreover, we found that the individuals who play critical roles in supporting the archaeological open source community are precariously employed workers. Contrary to popular claims about open source being inherently distributed, resilient, and open-ended, we found that, overall, archaeological open source is actually quite centralized, fragile, and based nearly exclusively on existing professional connections and endeavours.

### 2023-06-21 analysis/_02-background.qmd
Exclude old files from quarto renders
Add headings
Some more work on the intro and background

## What is open source?

Open source is a software development model that prioritizes transparent work processes. Initially driven by the idea that computer users should be free to understand and manipulate the software that they install on their computers (e.g., "free software", as initially conceived by the Free Software Foundation), open source has become a means of collaborative software development (Kelty 2008: chapter 3, especially starting at page 99). By putting one's code on the web without restriction on how it may be used or manipulated, this encourages creativity to flourish as people contribute to help improve the code base. Software thus emerges from the coordinated labour of worldwide volunteers, who shape the product according to the collective vision. An open code base may also be used to support alternative projects whose missions diverge from the original plan, and an entire project may be "forked", or taken in a new direction if contributors are dissatisfied with how core developers run things.

Open source has traditionally been referred to as being based on meritocratic principles. A good test of whether a contribution should be included in a published software release is whether it is functional. Moreover, with more eyes looking over a code base, it is easier to identify flaws with a contribution and flag potential bugs or security issues. This is all done in the spirit of producing functional code, and in ideal circumstances faulty contributions will be corrected before inclusion. Personal ego is minimized in favour of co-creating stable and functional outcomes. (Coleman 2012)

However, this is not the same as saying that open source is completely anarchic or based on the "wisdom of crowds". In fact, successful open source projects incorporate complex organizational structures, governance strategies, and forms or social mediation to help delegate and vet contributions made by distributed participants. They rely on, rather than eschew, institutional support structures, in order to motivate work, keep volunteer maintainers involved, and generally ensure that the project can be sustained over the long-term. Open source is more than just putting your code online; to be successful, it requires participation in a social experience. (Kelty 2008; O'Neill 2009; Ratto dissertation and derived works)

@@

## How does open source relate to open science?

The open science movement comprises a series of practices and principles intended to make research more accessible, transparent, and efficient. Although the concept of "open" is somewhat nebulous, in terms of its abstract definition and with regards to what real-world applications count as being open, one commonly-cited definition describes content that "can be freely used, modified, and shared by anyone for any purpose" (https://opendefinition.org/). This definition does not state what open is for, how to be open, or any sort of social or discursive framing behind the open movement. However, most open science advocates claim that they are motivated by a desire to facilitate novel research opportunities, make participation in scienctific research more equitable, reclaim science as a public good, and enhance how findings are validated and legitimized.

The idea that scientistis should generally contribute to a public domain of knowledge without profit motive has led to open science being heralded as revolutionary, community-oriented, and anti-capitalist means of production. However, while open science does have the _potential_ to effect radical change, this is not a given. The social and institutional contexts in which we do science is firmly embedded within capitalist and neoliberal power structures that reward individualistic competition and do little to actually encourage equitable and accessible research practices, and as such, make it difficult to fully embrace open science ideals. Moreover, the open science movement, which is dominated by STEM disciplines, prioritizes a grossly simplified and asocial notion of what science is and entails. Namely, it considers science as the accumulation and assembly of a species-level understanding of the world, which is not held by any one individual but is stored in seemingly value-neutral and disembodied media, facts and observations. This is manifested by digital telecommunications systems that host files, document processes, facilitate co-working opportunities, and perform automated processes. However, these systems have become so emblematic of open science that the use of these tools and resources designed to _support_ open science is often mistaken for _actually doing_ open science.

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain. But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (inlcuding: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs, and rarely anticipate use of their work beyond the limited scope of their primary targeted audience. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of tenure and promotion. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed and credit has been allocated, it is common for scientists to mark their projects as completed and move onto new endeavours. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly, and contributors may remain actively or sporadically involved to satisfy users' needs.

## Open archaeology at a crossroads

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a desire to facilitate novel research opportunities, to make participation in scienctific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. Open science advocates, including archaeologists, have put a lot of effort into developing information infrastructures, or systems designed to facilitate the transmission of data on a global scale. And as with open science in general, open archaeology has been wrapped up in discourse on the best tools to use, optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards (cf. Kansa and Kansa 2014; Kintigh 2014; Roosevelt 2015). This continues to be a prominent part of discourse on the development of open infrastructures (cf. Kansa 2022 on unique  identifiers; something about aDNA pipelines; Hacıgüzeller 2021 on CIDOC-CRM (though she is actually very critical); Huvila and his team on formal paradata), but more recent reflections on the use of these infrastructures recognize underlying and systemtic commitments and assumptions that inform how these systems are used and valued by various stakeholders. For instance, the pressures and circumstances of being an archaeologist, such as the need to publish stable research outputs, and archaeologists' tendency to move on from projects after they are completed, contribute towards shaping data sharing platforms as delegated archival services, whereupon the obligation to care for data over the long-term is passed on to others. In response to criticisms that highlight how these systems add little practical downstream value (Huggett 2018; Sobotkova 2018), open archaeology advocates have begun to prioritize the role that information infrastructures may play in contexts of data reuse, namely by implementing the FAIR principles in order to ensure that data are findable, accessible, interoperable, and reusable (Kansa 2023 in that recent special issue in Advances in Archaeological Practice; one or two additional sources). This example is matched by similar efforts to raise the value proposition of using open science tools and systems, such as through an emphasis on enhancing R package development and documentation skills so that code may be run and evaluated in alternative computing environments (cite Marwick and/or Roe), and [another example].

Additionally, archaeologists are becoming more comfortable showcasing and actively discussing undervalued research ouputs such as software and datasets, which are reviewed and published by venues like Peer Community In, the Journal of Archaeological Data, and the Journal of Open Source Software. Professional blogs, social media and podcasts are also gaining traction as proper venues for disseminating one's work. This is consistent with archaeology's historical willingness to experiment with new methods and media and its casual attitude towards scholarly communication. Archaeologists' embrace of methodological plurality and their positions as interdisciplinary scholars scattered across institutional boundaries has prepared them for this kind of pragmatic, improvisatory approach to scholarly communication via the web. There is enormous potential for archaeologists to reinvent novel forms of scholarly communication and extend a new set of collaborative commitments that go beyond those typically recognized by scholarly institutions, and which correspond with open science principles. But this depends on archaeologists' willingness to continue experimenting and on the community's willingness to buy into novel value regimes.

>>>>

# Open science and open source <!-- Background -->

## Open source

Open source is a software development model that prioritizes transparent work processes. Initially driven by the idea that computer users should be free to understand and manipulate the software that they install on their computers (e.g., "free software", as initially conceived by the Free Software Foundation), open source has become a means of collaborative software development (@kelty2008: chapter 3, especially starting at page 99). By putting one's code on the web without restriction on how it may be used or manipulated, this encourages creativity to flourish as people contribute to help improve the code base. Software thus emerges from the coordinated labour of worldwide volunteers, who shape the product according to the collective vision. An open code base may also be used to support alternative projects whose missions diverge from the original plan, and an entire project may be "forked", or taken in a new direction if contributors are dissatisfied with how core developers run things.

Open source has traditionally been referred to as being based on meritocratic principles. A good test of whether a contribution should be included in a published software release is whether it is functional. Moreover, with more eyes looking over a code base, it is easier to identify flaws with a contribution and flag potential bugs or security issues. This is all done in the spirit of producing functional code, and in ideal circumstances faulty contributions will be corrected before inclusion. Personal ego is minimized in favour of co-creating stable and functional outcomes (@coleman2012).

However, this is not the same as saying that open source is completely anarchic or based on the "wisdom of crowds". In fact, successful open source projects incorporate complex organizational structures, governance strategies, and forms or social mediation to help delegate and vet contributions made by distributed participants. They rely on, rather than eschew, institutional support structures, in order to motivate work, keep volunteer maintainers involved, and generally ensure that the project can be sustained over the long-term. Open source is more than just putting your code online; to be successful, it requires participation in a social experience (@kelty2008; Ratto dissertation and derived works).

@@

## Open science

The open science movement comprises a series of practices and principles intended to make research more accessible, transparent, and efficient. Although the concept of "open" is somewhat nebulous, in terms of its abstract definition and with regards to what real-world applications count as being open, one commonly-cited definition describes content that "can be freely used, modified, and shared by anyone for any purpose" (@openknowledgefoundation2023). This definition does not state what open is for, how to be open, or any sort of social or discursive framing behind the open movement. However, most open science advocates claim that they are motivated by a desire to facilitate novel research opportunities, make participation in scienctific research more equitable, reclaim science as a public good, and enhance how findings are validated and legitimized.

The idea that scientistis should generally contribute to a public domain of knowledge without profit motive has led to open science being heralded as revolutionary, community-oriented, and anti-capitalist means of production. However, while open science does have the _potential_ to effect radical change, this is not a given. The social and institutional contexts in which we do science is firmly embedded within capitalist and neoliberal power structures that reward individualistic competition and do little to actually encourage equitable and accessible research practices, and as such, make it difficult to fully embrace open science ideals (@mirowski2018). Moreover, the open science movement, which is dominated by STEM disciplines, prioritizes a grossly simplified and asocial notion of what science is and entails. Namely, it considers science as the accumulation and assembly of a species-level understanding of the world, which is not held by any one individual but is stored in seemingly value-neutral and disembodied media, facts and observations. This is manifested by digital telecommunications systems that host files, document processes, facilitate co-working opportunities, and perform automated processes. However, these systems have become so emblematic of open science that the use of these tools and resources designed to _support_ open science is often mistaken for _actually doing_ open science.

Open science is typically compared with the open source movement in that they both involve a distributed, digitally-mediated and worldwide labour force, who somehow derive rough consensus directed towards assets held in the public domain (cf. @tennant2020). But they differ in terms of the contexts in which they operate, the stakeholders involved, and the kinds of outcomes they produce. Whereas open source emerged from concern of consumer rights and then developed as a means of maintaining resilient and collectively motivated projects, open science comes out of a desire to make research practices more transparent and accessible. Open source is performed by professional and hobbyist software developers alike, and participants contribute in a wide variety of ways (inlcuding: programming, writing documentation, translating software and documentation, bug reporting, and financial support), but, in open science, scientists are usually the only participants actively involved in creating and maintaining contributions. Moreover, whereas open source projects often attract participants with varied stakes in the software and use cases in mind, open science projects are typically bounded by small communities of specialists with very particular needs, and rarely anticipate use of their work beyond the limited scope of their primary targeted audience. Additionally, open science is bounded by the professional contexts in which science operates, and as such, produces outputs that can be easily credited to specific sets of individuals for reasons of tenure and promotion. Open science projects whose contributions are supported by research funding also face sustainability concerns, as participants lose motivation to contribute once funding runs out. Once a project is completed and credit has been allocated, it is common for scientists to mark their projects as completed and move on to new endeavours. Open source projects, on the other hand, are motivated by a more practical need for the software to function properly, and contributors may remain actively or sporadically involved to satisfy users' needs.

The adoption of open source development models among archaeologists is generally informed by the broader open science movement, which is motivated by a genuine desire to facilitate novel research opportunities, to make participation in scienctific research more equitable, to reclaim science as a public good, and to enhance the means of validating findings. However, the predominant concern with implementing best tools to use, adopting optimal data processing pipelines, and tying into global, web-based infrastructures, protocols and standards (cf. @kansa2014; @kintigh2015; @roosevelt2015) distract from fundamental tensions and contradictions regarding the actual value of working in the open. For instance, @faniel2013 :299-301, @atici2013 :676-677, @huggett2018, @sobotkova2018, and @opitz2021 demonstrate that to make the reuse of archaeological data feasible and useful in a practical sense, it is necessary to re-introduce social friction that these infrastructures are designed to eliminate. In other words, the pressures and circumstances of being an archaeologist and doing archaeological research assert themselves when attempting to make practical use of these infrastructures, and therefore must be accounted for in their design and implementation. Here we aim to identify similar sources of dissonance with regards to the promise, potential, and actual implementation of open source software development models among archaeologists.
